Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Starting a chapter on deadlock prevention

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapters~\ref{ch:cxlvalues} and~\ref{ch:cxlmachine} present some more
details of the CXL language and machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section.
\item Chapter~\ref{ch:naive} presents various flawed implementations
of critical sections.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to the mutual exclusion problem.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for mutual exclusion.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution for the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate over an unreliable network.
\end{itemize}

@@

>>>>

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapters~\ref{ch:cxlvalues} and~\ref{ch:cxlmachine} present some more
details of the CXL language and machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section, and Chapter~\ref{ch:naive} presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant implementation of a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution applied to the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\chapter{Deadlock Prevention}
\label{ch:deadlockprevention}

As introduced in Chapter~\ref{ch:deadlock}, deadlock is the situation when there are
a collection of processes waiting for one another to release one or more resources.
There are four conditions that must hold for deadlock to occur~\cite{CES71}:
\begin{enumerate}
\item \emph{Mutual Exclusion}: each resource can only be used by one process at a time:
\item \emph{Hold and Wait}: each process holds resources it already allocated while it
waits for other resources that it needs;
\item \emph{No Preemption}: Resources cannot be forcibly taken back from processes that
allocated them;
\item \emph{Circular Wait}: There exists a directed circular chain of processes, each waiting
to allocate a resource held by the next.
\end{enumerate}

Preventing deadlock thus means preventing that one of these conditions occurs.
However, mutual exclusion is not easily prevented (although, for some resources it is
possible, as demonstrated in Chapter~\ref{ch:nonblocking}).
Havender~\cite{Havender68} proposed the following techniques that avoid the remaining
three conditions:

\begin{itemize}
\item \emph{No Hold and Wait}: a process must request all resources it is going to
need at the same time;
\item \emph{Preemption}: if a process is denied a request for a resource, it must
release all resources that it has already acquired and start over;
\item \emph{No Circular Wait}: define an ordering on all resources and allocate
resources in a particular order.
\end{itemize}

We suggested the last technique to address the deadlock issue in the Dining
Philosophers problem (Chapter~\ref{ch:deadlock}).
